---
title: "figures"
author: "Alexander Haglund"
date: "2022-11-10"
output:
  html_notebook: default
  workflowr::wflow_html:
    toc: true
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---
### libraries
```{r, message=FALSE}
library(ggplot2)
library(viridis)
library(ggsci)
library(dplyr)
library(cowplot)
library(grid)
library(tidyr)
suppressMessages(library(reshape))
color_pal=ggsci::pal_nejm("default")(8)
colorvec<-c(Astrocytes=color_pal[1],
           Endothelial=color_pal[2],
           Excitatory=color_pal[3],
           Inhibitory=color_pal[4],Microglia=color_pal[5],
           ODC=color_pal[6],OPC=color_pal[7],Pericytes=color_pal[8])
```

# FIGURE 1

coming soon

# FIGURE 2

### Figure 2a
```{r}
coloc<-read.table("data/COLOC_MR_RESULTS/2022-10-25_FULL_COLOC_RES.txt")
x<-coloc[coloc$GWAS %in% "AD",]



genes<-x[x$PP.H4.abf>0.5,]$gene
x<-x[x$gene %in% genes,]
x$gene<-factor(x$gene,levels=unique(x$gene))

g<-ggplot(x,aes(x=celltype,y=gene,fill=PP.H4.abf))
g<-g+geom_tile(aes(fill=round(PP.H4.abf,2)),colour="black")+
geom_text(aes(label = round(PP.H4.abf, 2)),size=5*0.36,family="Helvetica")+
scale_fill_viridis()+
theme_classic()+
scale_y_discrete(limits=rev,expand = c(0, 0))+
scale_x_discrete(expand = c(0, 0),position="top")+
xlab("Alzheimer's Disease")
g
```

### Figure 2b
coming soon

### Figure 2c
coming soon

### Figure 2d
```{r}

coloc_results<-read.table("data/COLOC_MR_RESULTS/2022-10-25_FULL_COLOC_RES.txt")
fig_dir<-"FIGURES/Figure_2/"

coloc_results<-coloc_results[coloc_results$PP.H4.abf>0.5,]


g<-ggplot(coloc_results,aes(y=GWAS))+geom_bar(color="black",fill="#166FA2")+
geom_text(aes(label=..count..),stat="count",hjust=-0.8,size=5/(14/5))+
labs(x="Number of colocalisations")+
theme_classic()+scale_x_continuous(limits=c(0,120),expand = c(0, 0))+
theme(text=element_text(family="Helvetica",face="bold"),
                                             axis.text.x=element_text(size=5,face="bold"),
                                            axis.text.y=element_text(size=5),
     axis.title.x=element_text(size=5),axis.title.y=element_text(size=5))
g
```

### Figure 2e
```{r}
coloc<-read.table("data/COLOC_MR_RESULTS/2022-10-25_FULL_COLOC_RES.txt")
##rename SCV caudate for spacing
coloc_results$GWAS<-sapply(coloc_results$GWAS,function(x){
    if(x=="SCV.CAUDATE"){
        x<-"SCV"}
    return(x)})

##filter to keep coloc hits
coloc_results_filtered<-coloc_results[coloc_results$PP.H4.abf>0.5,]


gwas_list<-unique(coloc_results_filtered$GWAS)
reslist<-list()

##count number of cell types per trait
for(i in 1:length(gwas_list)){
    tmp<-coloc_results_filtered %>% filter(GWAS==gwas_list[i])
    freq_table<-as.data.frame(table(tmp$celltype))
    tmp$celltype_freq<-freq_table[match(tmp$celltype,freq_table$Var1),]$Freq

    #scale by total number of colocs (otherwise larger gwases will have really large points)
    # tmp$celltype_freq<-tmp$celltype_freq/nrow(tmp)
    reslist[[i]]<-tmp
    }
coloc_results_filtered<-as.data.frame(do.call(rbind,reslist))

color_pal=ggsci::pal_nejm("default")(8)
g<-ggplot(coloc_results_filtered,aes(y=GWAS,x=celltype,size=celltype_freq,fill=celltype))+
geom_point(pch=21)+
scale_size(range=c(1,3))+
# geom_text(aes(label=celltype_freq),size=5/(14/5),vjust=-1.1)+
theme_classic()+
scale_fill_manual(values=color_pal)+
theme(text=element_text(family="Helvetica",face="bold"),
                                             axis.text.x=element_text(size=5,angle=45),
                                            axis.text.y=element_text(size=5),
     axis.title.x=element_text(size=5),axis.title.y=element_text(size=5),
      legend.text=element_text(size=5),legend.title=element_text(size=5),
      legend.spacing.y = unit(0.05, 'cm'),legend.position = "none")
g

```

# FIGURE 3

The code presented here is in step 5 (Epigenetic Intersection) of the downstream results section.
The basic syntax is a geom_tile() function in ggplot with colours matching cell-types. More in data/MARKDOWN/helper_funcs.r

This is only for Fig 3.a - Fig 3.c and Fig 3.d were made using the UCSC track browser.

```{r}

indir<-"data/EXT_DATASETS//RESULTS/"
fig_dir<-"FIGURES/Figure_3/"

g1<-readRDS(paste0(indir,"oligo_intersect_ggobject.rds"))
g2<-readRDS(paste0(indir,"Excneuron_intersect_ggobject.rds"))
g3<-readRDS(paste0(indir,"Inneuron_intersect_ggobject.rds"))
g4<-readRDS(paste0(indir,"microglia_intersect_ggobject.rds"))

g1<-g1+theme(legend.position = "none")
g2<-g2+theme(legend.position = "none")
g3<-g3+theme(legend.position = "none")
g4<-g4+theme(legend.position = "none")

```

## Oligo
```{r}
g1
```
## ExcNeur
```{r}
g2
```
## InNeur
```{r}
g3
```
## Microglia
```{r}
g4
```
# FIGURE 4
## Figure 4a
bind data together
```{r}
#bind data together

pqtl<-read.table("data/TABLES/pQTL_table.txt")
stitch<-read.table("data/TABLES//stitch_table.txt")
dgidb<-read.table("data/TABLES/dgidb_table.txt")
opentargets<-read.table("data/TABLES/OpenTargets_table.txt")
coloc<-read.table("data/COLOC_MR_RESULTS//2022-10-25_FULL_COLOC_RES.txt")
full<-read.table("data/COLOC_MR_RESULTS//2022-10-25_FULL_MR_RES.txt")
full<-full[full$IVW<0.05,]
full$trait_gene<-paste0(full$GWAS,"_",full$gene)
full$IVW_dir<-sapply(full$IVW_beta,function(x){
    if(x>0){
        return("positive")
        }else{
        return("negative")
        }
    })
direction_vector<-vector()
for(i in 1:nrow(full)){
    tmp<-full[full$trait_gene %in% full$trait_gene[i],]
    betas<-tmp$IVW_beta
    
    if(all(betas>0)==TRUE){
    dir<-"positive"
    } else if(all(betas<0)==TRUE){
    dir<-"negative"
    }else{
        dir<-"N/A"
        }
    direction_vector<-c(direction_vector,dir)    
        }
full$IVW_dir<-direction_vector    


coloc$trait_gene_ct<-paste0(coloc$GWAS,"_",coloc$gene,"_",coloc$celltype)
trait_gene_ct<-paste0(pqtl$GWAS,"_",pqtl$gene,"_",pqtl$celltype)
coloc<-coloc[match(trait_gene_ct,coloc$trait_gene_ct),]

plot_df<-data.frame(gene_trait=paste0(pqtl$GWAS,".",pqtl$gene),
                    pQTL=pqtl$pQTL_hit,
                    STITCH=stitch$STITCH_intersect,
                    DGidb=dgidb$DGIDB_intersect,
                    OpenTargets=opentargets$OpenTargets_disease_hit,
                   coloc=coloc$PP.H4.abf,
                   celltype=pqtl$celltype,
                   IVW_dir=full$IVW_dir)

```

prep plots
```{r}

##intersection plot
plot_df1<-plot_df[,c("gene_trait","pQTL","STITCH","DGidb","OpenTargets")]
melt_df<-melt(plot_df1,id=c("gene_trait"))
ggplot_mainplot1<-ggplot(melt_df,aes(x=variable,y=gene_trait,fill=value))+geom_tile(color="black")+
scale_fill_manual(values=c("#c8c8c8","#A6CEE3","#1F78B4"))+scale_x_discrete(expand=c(0,0))

##direction
ivw_dir<-plot_df[,c("gene_trait","IVW_dir")]
ivw_dir$trait<-"gwas"
ivw_dir1<-ivw_dir
ggplot_ivwdir1<-ggplot(ivw_dir1,aes(y=gene_trait,x=trait,fill=IVW_dir,group=IVW_dir))+geom_tile(colour="black")+
scale_fill_manual(values = c("#c8c8c8","#F8766D","#00BFC4"),guide = guide_legend(reverse = TRUE))+
theme_classic()

##celltypes

celltype_bar<-plot_df[,c("gene_trait","celltype")]
celltype_bar1<-celltype_bar
test<-celltype_bar1 %>% 
count(gene_trait,celltype,name="count") %>% 
complete(gene_trait,celltype)
newvec<-vector()
for(i in 1:nrow(test)){
    if(is.na(test$count[i])){
        newvec<-c(newvec,NA)
        }else{
        newvec<-c(newvec,test$celltype[i])
        }
    }
test$new<-newvec
# melt_df<-melt(celltype_bar1,id=c("gene_trait"))
# df2 = celltype_bar1 %>% complete(gene_trait,celltype)
ggplot_celltype1<-ggplot(test,aes(x=celltype,y=gene_trait,fill=new))+geom_tile(color="black")+
scale_fill_manual(values=colorvec,na.value="white")

```

### plot1
```{r}
ggplot_mainplot1<-ggplot_mainplot1+theme_classic()+
theme(axis.text.x=element_text(size=5,face="bold",angle=45,vjust=0.8,hjust=0.8,family="Helvetica"),
      axis.text.y=element_text(size=5,face="bold",vjust=0.8,family="Helvetica"),axis.title.y=element_blank(),
      axis.line.y=element_blank(),axis.line.x=element_blank(),
      axis.title.x=element_blank(),legend.key.size=unit(0.7,"cm"),
      legend.title=element_blank(),
      legend.text=element_text(size=5),
      panel.background = element_rect(fill='transparent'),
      plot.background = element_rect(fill='transparent', color=NA),
      panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
      legend.background = element_rect(fill='transparent'),
      legend.key.height= unit(0.3, 'cm'),
      legend.key.width= unit(0.2, 'cm'))#

ggplot_ivwdir1<-ggplot_ivwdir1+theme_classic()+
theme(axis.text.x=element_blank(),
      axis.text.y=element_blank(),axis.title.y=element_blank(),axis.ticks.y=element_blank(),
      axis.ticks.x=element_blank(),
      axis.line.y=element_blank(),axis.line.x=element_blank(),
      axis.title.x=element_blank(),legend.key.size=unit(0.7,"cm"),
      legend.text=element_text(size=5), legend.title=element_text(size=5),
      panel.background = element_rect(fill='transparent'),
      plot.background = element_rect(fill='transparent', color=NA),
      legend.background = element_rect(fill='transparent'),
      legend.key.height= unit(0.3, 'cm'),
      legend.key.width= unit(0.2, 'cm'))#


ggplot_celltype1<-ggplot_celltype1+theme_classic()+
theme(axis.text.x=element_text(size=5,face="bold",angle=45,vjust=0.8,hjust=0.8,family="Helvetica"),
      axis.text.y=element_blank(),axis.title.y=element_blank(),axis.ticks.y=element_blank(),
      axis.line.y=element_blank(),axis.line.x=element_blank(),
      axis.title.x=element_blank(),legend.key.size=unit(0.7,"cm"),
      legend.title=element_blank(),
      legend.text=element_text(size=5),
      panel.background = element_rect(fill='transparent'),
      plot.background = element_rect(fill='transparent', color=NA),
      legend.background = element_rect(fill='transparent'),
      legend.key.height= unit(0.3, 'cm'),
      legend.key.width= unit(0.2, 'cm'))#

ggplot_mainplot1_legend<-get_legend(ggplot_mainplot1)
celltype_legend<-get_legend(ggplot_celltype1)
ivw_dir<-get_legend(ggplot_ivwdir1)


ggplot_ivwdir1<-ggplot_ivwdir1+theme(plot.margin = unit(c(0,0,0,0), "cm"),legend.position = "none")
ggplot_mainplot1<-ggplot_mainplot1+theme(plot.margin = unit(c(0,0,0,0), "cm"),legend.position = "none")
ggplot_celltype1<-ggplot_celltype1+theme(plot.margin = unit(c(0,0,0,0), "cm"),legend.position="none")

g<-plot_grid(ggplot_mainplot1,ggplot_celltype1,ggplot_ivwdir1, align = "h", ncol =3, rel_widths = c(0.008,0.005,0.001))
g
```


# FIGURE 5
## Figure 5a
```{r}
full<-read.table("data/COLOC_MR_RESULTS//2022-10-25_FULL_MR_RES.txt")
full<-full[full$IVW<0.05,]
full$trait_gene<-paste0(full$GWAS,"_",full$gene)
full$IVW_dir<-sapply(full$IVW_beta,function(x){
    if(x>0){
        return("CAUSAL")
        }else{
        return("PROTECTIVE")
        }
    })

tmp<-full[full$GWAS=="AD",]
tmp<-tmp[order(tmp$celltype),]
tmp$gene<-factor(tmp$gene,levels=unique(tmp$gene))
color_pal=ggsci::pal_nejm("default")(8)
colorvec<-c(Astrocytes=color_pal[1],
           Endothelial=color_pal[2],
           Excitatory=color_pal[3],
           Inhibitory=color_pal[4],Microglia=color_pal[5],
           ODC=color_pal[6],OPC=color_pal[7],Pericytes=color_pal[8])

g<-ggplot(tmp,aes(y=gene,x=IVW_beta,fill=celltype))+
geom_point(colour="black",shape=21,stroke=0.3,size=2)+
scale_fill_manual(values = colorvec)+
theme_classic()+
geom_vline(xintercept=0,linetype="dotted")+scale_x_continuous(limits=c(-0.6,0.6))
g2<-g+
theme(text=element_text(family="Helvetica",size=5),legend.text = element_text(family="Helvetica",size=5,face="bold"),
      legend.title=element_blank(),
      legend.spacing.y = unit(-1, 'cm'),axis.text.y=element_text(family="Helvetica",size=5,face="bold"))+guides(fill = guide_legend(byrow = TRUE))
g2
```

## Figure 5b

```{r}
full<-read.table("data/COLOC_MR_RESULTS//2022-10-25_FULL_MR_RES.txt")
plot_df<-data.frame()
for(i in 1:length(unique(full$GWAS))){
    gwas<-unique(full$GWAS)[i]
    tmp_full<-full[full$GWAS==gwas,]
    cellvec<-as.vector(table(tmp_full$celltype))
    plot_df<-rbind(plot_df,data.frame(gwas=gwas,
                                      celltype=names(table(tmp_full$celltype)),
                                      occurrence=cellvec))
    }
g<-ggplot(data=plot_df,aes(y=occurrence,x=gwas,fill=celltype))+
geom_bar(stat="identity",colour="black",size=0.25)+
scale_fill_manual(values=color_pal)+theme_classic()+
geom_text(aes(label=occurrence),family="Helvetica",size=5*0.36, position = position_stack(vjust = 0.5))+
theme_classic()+scale_y_continuous(expand=c(0,0))+
labs(title="Cell-type proportions in MR results",size=7,family="Helvetica")+
theme(text=element_text(size=7,family="Helvetica"),axis.text.x=element_text(size=5,face="bold",angle=90),
                                  axis.text.y=element_text(size=5),
                                  axis.title.y=element_blank(),
                                 axis.title.x=element_blank(),legend.position="none",
     title=element_text(size=7,face="bold"))
g
```
## Figure 5c
```{r}
full<-read.table("data/COLOC_MR_RESULTS//2022-10-25_FULL_MR_RES.txt")
full$celltype_gene<-paste0(full$celltype,".",full$gene)
scz<-full[full$GWAS=="SCZ",]
iq<-full[full$GWAS=="IQ",]
common<-intersect(iq$celltype_gene,scz$celltype_gene)

##remove ct/gene combinations in other traits - we are only interested in the SCZ/IQ overlap
# filtered<-full[full$celltype_gene %in% common,]
# filtered<-filtered[!filtered$GWAS %in% c("SCZ","IQ"),]
# to_exclude<-unique(filtered$celltype_gene)
# common<-common[!common %in% to_exclude]



#filter both
scz<-scz[match(common,scz$celltype_gene),]
iq<-iq[match(common,iq$celltype_gene),]
df<-data.frame(celltype_gene=common,IQ=iq$IVW_beta,SCZ=scz$IVW_beta)


##now plot
ylims=c(-0.2,0.2)
xlims=c(-0.7,0.7)

g<-ggplot(df,aes(x=SCZ,y=IQ,fill=celltype_gene))+
geom_point(shape=21,size=2)+
scale_y_continuous(limits=ylims)+
scale_x_continuous(limits=xlims)+
geom_vline(xintercept =0,linetype="dotted")+
geom_hline(yintercept =0,linetype="dotted")+theme_classic()+
theme(text=element_text(size=5,family="Helvetica"),legend.title=element_blank(),legend.text = element_text(family="Helvetica",size=5,face="bold"),legend.spacing.y = unit(-1, 'cm'))+
xlab("SCZ - MR beta")+ylab("IQ - MR beta")+
guides(fill = guide_legend(byrow = TRUE))
g
```

## Figure 5d

```{r}
full<-read.table("data/COLOC_MR_RESULTS//2022-10-25_FULL_MR_RES.txt")
full$celltype_gene<-paste0(full$celltype,".",full$gene)
scz<-full[full$GWAS=="SCZ",]
neur<-full[full$GWAS=="NEUR",]
common<-intersect(neur$celltype_gene,scz$celltype_gene)

# filtered<-full[full$celltype_gene %in% common,]
# filtered<-filtered[!filtered$GWAS %in% c("SCZ","NEUR"),]
# to_exclude<-unique(filtered$celltype_gene)
# common<-common[!common %in% to_exclude]


#filter both
scz<-scz[match(common,scz$celltype_gene),]
neur<-neur[match(common,neur$celltype_gene),]
df<-data.frame(celltype_gene=common,NEUR=neur$IVW_beta,SCZ=scz$IVW_beta)


##now plot
ylims=c(-0.2,0.2)
xlims=c(-0.8,0.8)

g<-ggplot(df,aes(x=SCZ,y=NEUR,fill=celltype_gene))+
geom_point(shape=21,size=2)+
scale_y_continuous(limits=ylims)+
scale_x_continuous(limits=xlims)+
geom_vline(xintercept =0,linetype="dotted")+
geom_hline(yintercept =0,linetype="dotted")+theme_classic()+
theme(text=element_text(size=5,family="Helvetica"),legend.text = element_text(family="Helvetica",size=5,face="bold"),legend.title=element_blank(),legend.spacing.y = unit(-1, 'cm'))+
xlab("SCZ - MR beta")+ylab("NEUR - MR beta")+guides(fill = guide_legend(byrow = TRUE))
g
```